---
title: "beanz: Bayesian Analysis of Heterogeneous Treatment Effect"
author: "Chenguang Wang and Ravi Varadhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{beanz: Bayesian Analysis of Heterogeneous Treatment Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, eval=T, echo=FALSE}
require(beanz);
```

# Introduction

In patient-centered outcomes research, it is vital to assess the heterogeneity
of treatment effects (HTE) when making health care decisions for an individual
patient or a group of patients. Nevertheless, it remains challenging to evaluate
HTE based on information collected from clinical studies that are often designed
and conducted to evaluate the efficacy of a treatment for the overall
population. The Bayesian framework offers a principled and flexible approach to
estimate and compare treatment effects across subgroups of patients defined by
their characteristics.

*R* package **beanz** provides functions to facilitate the conduct of Bayesian
analysis of HTE and a web-based graphical user interface for users to conduct
such Bayesian analysis in an interactive and user-friendly manner.

# Data accepted by **beanz**

There are two types of data structures that **beanz** recognizes:

- Summary treatment effect data: Each row should represent a subgroup with
covariates that define the subgroup, estimated treatment effect in the subgroup
and variance for the estimation.

- Patient level raw data: Each row should represent a patient with covariates
that define the subgroup in which the patient belongs to, treatment indicator
and outcome. The outcome can be binary, continuous, or time to event.

The **beanz** package provides dataset *solvd.sub* from the *SOLVD* trial as an
example *Patient level raw data* dataset.

# Estimate subgroup effect

If *Patient level raw data* is provided, the package provides function
*r.get.subgrp.raw* for estimating subgroup effect for each subgroup. The return
value from *r.get.subgrp.raw* is a data frame with the format of *Summary
treatment effect data*.

The example is as follows:

```{r, eval=T, echo=TRUE}

var.cov    <- c("sodium", "lvef", "any.vasodilator.use");
var.resp   <- "y";
var.trt    <- "trt";
var.censor <- "censor";
resptype   <- "survival";

subgrp.effect <- r.get.subgrp.raw(solvd.sub,
                                  var.resp   = var.resp,
                                  var.trt    = var.trt,
                                  var.cov    = var.cov,
                                  var.censor = var.censor,
                                  resptype   = resptype);
print(subgrp.effect);

```

# Bayesian HTE models

The function *call.stan* calls *rstan::sampling* to draw samples for different
Bayesian models. The following models are available in the current version of
**beanz**:

- *nse*: No subgroup effect model
- *fs*: Full stratification model
- *sr*: Simple regression model
- *bs*: Basic shrinkage model
- *srs*: Simple regression with shrinkage model
- *ds*: Dixon-Simon model
- *eds*: Extended Dixon-Simon model.


The following examples show how *No subgroup effect model* (*nse),
*Simple regression model* (*sr*) and *Basic shrinkage model* (*bs*) are called:

```{r, eval=T, echo=TRUE}

var.cov    <- c("sodium", "lvef", "any.vasodilator.use");
var.estvar <- c("Estimate", "Variance");

rst.nse <- call.stan("nse", dat.sub=subgrp.effect,
                     var.estvar = var.estvar, var.cov = var.cov,
                     lst.par.pri = list(vtau=1000, vrange=c(0,0)),
                     chains=1, iter=4000,
                     warmup=2000, thin=2, seed=1000);

rst.sr  <- call.stan("sr", dat.sub=subgrp.effect,
                     var.estvar = var.estvar, var.cov = var.cov,
                     lst.par.pri = list(vtau=1000, vgamma=1000, vrange=c(0,0)),
                     chains=1, iter=4000,
                     warmup=2000, thin=2, seed=1000);

rst.bs  <- call.stan("bs", dat.sub=subgrp.effect,
                     var.estvar = var.estvar, var.cov = var.cov,
                     lst.par.pri = list(vtau=1000, vw=100, vrange=c(-0.1,0.1)),
                     chains=1, iter=4000, warmup=2000, thin=2, seed=1000);

```

# Results presentation

## Posterior subgroup treatment effect summary

Posterior subgroup treatment effect can be summarized and presented by functions
*r.summary.stan*, *r.plot.stan* and *r.forest.stan*. These functions allows to
include a subgroup from another model (i.e. No subgroup effect model) as a
reference in the results.

### Simple regression model
```{r, eval=T, echo=TRUE, fig.width=6, fig.height=5}
sel.grps <- c(1,4,5);
tbl.sub <- r.summary.stan(rst.sr, ref.stan.rst=rst.nse, ref.sel.grps=1);
print(tbl.sub);
r.plot.stan(rst.sr, sel.grps = sel.grps, ref.stan.rst=rst.nse, ref.sel.grps=1);
r.forest.stan(rst.sr, sel.grps = sel.grps, ref.stan.rst=rst.nse, ref.sel.grps=1);
```

### Basic shrinkage model
```{r, eval=T, echo=TRUE, fig.width=6, fig.height=5}
tbl.sub <- r.summary.stan(rst.bs, ref.stan.rst=rst.nse, ref.sel.grps=1);
print(tbl.sub);
r.plot.stan(rst.bs, sel.grps = sel.grps, ref.stan.rst=rst.nse, ref.sel.grps=1);
r.forest.stan(rst.bs, sel.grps = sel.grps, ref.stan.rst=rst.nse, ref.sel.grps=1);
```

## Posterior subgroup treatment effect comparison

Posterior subgroup treatment effect can be compared between subgroups by functions
*r.summary.comp*, *r.plot.comp* and *r.forest.comp*.

### Simple regression model
```{r, eval=T, echo=TRUE, fig.width=6, fig.height=5}
tbl.sub <- r.summary.comp(rst.sr, sel.grps=sel.grps);
print(tbl.sub);
r.plot.stan(rst.sr, sel.grps = sel.grps);
r.forest.stan(rst.sr, sel.grps = sel.grps);
```

### Basic shrinkage model
```{r, eval=T, echo=TRUE, fig.width=6, fig.height=5}
tbl.sub <- r.summary.comp(rst.bs, sel.grps=sel.grps);
print(tbl.sub);
r.plot.comp(rst.bs, sel.grps = sel.grps);
r.forest.comp(rst.bs, sel.grps = sel.grps);
```

## Overall summary
**beanz** provides function *r.rpt.tbl* to generate the summary posterior
subgroup treatment effect table from the model selected by DIC (i.e. the model
with the smallest DIC):

```{r, echo=TRUE}
lst.rst     <- list(nse=rst.nse, sr=rst.sr, bs=rst.bs);
tbl.summary <- r.rpt.tbl(lst.rst, dat.sub = subgrp.effect, var.cov = var.cov);
print(tbl.summary);
```


# Graphical User Interface

With package *shiny* installed, *beaz* provides a web-based graphical user
interface (GUI) for conducting the HTE analysis in an user-friendly interactive
manner. The GUI can be started by

```{r, eval=F}
run.beanz();
```

# Toolbox

Package **beanz** provides function *r.gailsimon* that implements the Gail-Simon
test for qualitative interactions:

```{r, echo=T}
gs.pval <- r.gailsimon(subgrp.effect$Estimate,
                       sqrt(subgrp.effect$Variance));
print(gs.pval);
```

The result show that there is no significant qualitative interactions according
to the Gail-Simon test.
